### 2016-12-14 paper/1_introduction.tex
Refactor format of paper and some graphics

Analysis tools traditionally do not scale well for vast data sets.
Because of the exponential data growth, analytic capacities have to improve at a similar speed to keep analyses feasible. 
Usually, one tries to speed up computers by vertical or horizontal scaling.
To scale vertically means that we add more resources to a single computer.
For example, the main memory size or the CPU frequency of a computer can be increased to improve computational power.
In contrast to that, horizontal scaling means to add more computer nodes to a system.
By having more than one node, the work can be split up and distributed across the nodes.

The emerging multi-core and distributed systems pose new challenges for programmers.
Since they have to be able to reason about interwoven parallel control flows, parallel program development is highly cumbersome and error-prone.
Therefore, new programming models are conceived, a development that relieves the programmer from the tedious low-level tasks related to parallelization such as load-balancing, scheduling of parallel tasks, and fault recovery.
With these new models, programmers can concentrate on the actual algorithm and use case rather than reasoning about distributed systems. 
These are the reasons why Google's MapReduce~\cite{dean:c2008a} framework and its open source re-implementation Hadoop~\cite{hadoop:2008a} became so popular among scientists as well as engineers.

However, MapReduce and other frameworks force the user to express the program in a certain way, which is often not natural or intuitive to a user coming from a different domain.
Especially, in the field of data analytics and machine learning programs are usually expressed in a mathematical form.
Therefore, systems such as MATLAB~\cite{matlab} and R~\cite{r:1993a} are widely used and recognized for their fast prototyping capabilities and their extensive mathematical libraries.
However, these linear algebra systems lack proper support for automatic parallelization on large clusters and are thus restricting the user to a single workstation.
Therefore, the amount of processable data is limited to the size of the main memory, which constitutes a serious drawback for real-world applications.

\begin{figure}[t!]
\centering
\includegraphics[height=0.175\paperheight]{images/systemArchitecture.png}
\caption{The layered system architecture of Gilbert consisting of the language, intermediate and runtime layer.}
\label{fig:systemArchitecture}
\end{figure}

As a solution we propose Gilbert, a distributed sparse linear algebra environment whose name is a homage to William Gilbert Strang.
Gilbert provides a MATLAB-like language for distributed sparse linear algebra operations. It has a layered architecture, which is shown in Figure \ref{fig:systemArchitecture}. The first layer is the language layer. It contains all functionality to parse the given Gilbert code and to compile it into an intermediate representation (IR). The second layer is the intermediate layer and it receives the IR of the Gilbert code. The intermediate format is the ideal representation to apply language independent high-level transformations. The Gilbert optimizer applies several algebraic optimizations prior to the generation of the execution plan. The execution plan generator translates the optimized IR into a specific execution plan, depending on the selected execution engine. Once the program has been translated into an execution engine's specific plan, it is executed on the respective back end.

Our main contributions are:
\begin{itemize}
 \item Gilbert allows to write MATLAB-like code for sparse linear algebra and execute it on massively parallel dataflow systems.
 \item Its expressiveness allows to quickly develop scalable algorithms to analyze web-scale data.
 \item We introduce a novel fixed-point operator which replaces loops by recursion and can be efficiently parallelized.
 \item Furthermore, we demonstrate that dataflow optimizers can be used to automatically select a good matrix multiplication strategy. 
\end{itemize}

>>>>

Analysis tools traditionally do not scale well for vast data sets.
Because of the exponential data growth, analytic capacities have to improve at a similar speed.
This problem can be tackled by vertical and horizontal scaling of computers.
To scale vertically means that we add more resources to a single computer.
For example, the main memory size or the CPU frequency of a computer can be increased to improve computational power.
In contrast to that, horizontal scaling refers to adding more computer nodes to a system.
By having more than one node, computational work can be split up and distributed across multiple computers.

The emerging field of multi-core and distributed systems poses new challenges for programmers.
Since they have to be able to reason about interwoven parallel control flows, parallel program development is highly cumbersome and error-prone.
Therefore, new programming models are conceived, a development that relieves the programmer from tedious low-level tasks related to parallelization such as load-balancing, scheduling, and fault tolerance.
With these new models, programmers can concentrate on the actual algorithm and use case rather than reasoning about distributed systems. 
These are the reasons why Google's MapReduce~\cite{dean:c2008a} framework and its open source re-implementation Hadoop~\cite{hadoop:2008a} became so popular among scientists as well as engineers.

MapReduce and other frameworks, however, force users to express programs in a certain way which is often not natural or intuitive to users from different domains.
Especially, in the field of data analytics and machine learning programs are usually expressed in a mathematical form.
Therefore, systems such as MATLAB~\cite{matlab} and R~\cite{r:1993a} are widely used and recognized for their fast prototyping capabilities and their extensive mathematical libraries.
However, these linear algebra systems lack proper support for automatic parallelization on large clusters and are thus restricting the user to a single workstation.
Therefore, the amount of processable data is limited to the size of a single machine's memory, which constitutes a serious drawback for real-world applications.

\begin{figure}[!h]
\centering
\includegraphics[height=0.15\paperheight]{images/systemArchitecture2.png}
\caption{Gilbert's system architecture consisting of the language, intermediate and runtime layer.}
\label{fig:systemArchitecture}
\end{figure}

As a solution we propose Gilbert, a distributed sparse linear algebra environment whose name is a homage to William Gilbert Strang.
Gilbert provides a MATLAB-like language for distributed sparse linear algebra operations. 
It has a layered architecture, which is shown in Figure \ref{fig:systemArchitecture}. 
The first layer is the language layer. 
This layer parses Gilbert code and compiles it into an intermediate representation (IR). 
The second layer is the intermediate layer, which is given the IR. 
The intermediate format is the ideal representation to apply language independent high-level transformations. 
The Gilbert optimizer applies several algebraic optimizations prior to the generation of the execution plan. 
The execution plan generator translates the optimized IR into an execution engine specific plan.
This execution plan is then executed on the respective back end.

Our main contributions are:

\begin{itemize}
 \item Gilbert allows to write MATLAB-like code for sparse linear algebra and execute it on massively parallel dataflow systems.
 \item Its expressiveness allows to quickly develop scalable algorithms to analyze web-scale data.
 \item We introduce a novel fixed-point operator which replaces loops by recursion and can be efficiently parallelized.
 \item Furthermore, we demonstrate that dataflow optimizers can be used to automatically select a good matrix multiplication strategy. 
\end{itemize}

### 2016-12-14 paper/3_gilbert_features.tex
Refactor format of paper and some graphics

Gilbert's language has been strongly inspired by MATLAB, which should ease its adoption.
The language is a fully functional subset of MATLAB and its grammar is specified in \cite{Rohrmann2014}.
The elementary data type is the matrix. 
Gilbert supports arbitrary $2$-dimensional matrices whose elements can be \emph{double} or \emph{boolean}.
Vectors are not represented by a special type but instead are treated as a single column/row matrix.
Additionally, scalar \emph{double}, \emph{boolean}, and \emph{string} values are supported. 
Gilbert also implements cell array types.
A cell array is defined using curly braces and commas to separate individual values. 
The cells can be accessed by an index appended in curly braces to the cell array variable. 
\Cref{lst:cellArray} shows how to define and access them.
\begin{listing}[!h]
    \begin{lstlisting}[language=Matlab,
        commentstyle=\color{black},
        stringstyle=\color{black},
        xleftmargin=.2\textwidth,
        basicstyle=\footnotesize
    ]
c = {true, 2*2, 'cell', 'array'};
b = c{1} & false; % = false
d = c{2} ^ 2; % = 16
s = {c{4}, c{3}}; % = {'array', 'cell'} 
    \end{lstlisting}
  \caption{Cell array usage in Gilbert. Definition of a 4 element cell array, which is accessed subsequently.}
  \label{lst:cellArray}
\end{listing}
Gilbert supports the basic linear algebra operations defined on matrices and scalars. 
They include among others \code{+}, \code{-}, \code{/}, and \code{*}, whereas \code{*} denotes the matrix-matrix multiplication and all other operations are interpreted cell-wise. 
The cell-wise multiplication is indicated by a point preceding the operator. 
Gilbert also supports comparisons operators such as \code{>}, \code{>=}, \code{==}, and \code{\textasciitilde=}. 
Besides the basic arithmetic operations, the user can also define named functions and anonymous functions. 
The syntax of anonymous functions adheres to the MATLAB syntax: \code{@(x,y) x*x + y*y}. 
An important aspect of MATLAB are loops. 
MATLAB permits the user to express \code{for} and \code{while} loops. 
However, these loops can have side effects. 
The problem of parallelization of iterations with side effects is that the referenced external state has to be maintained. 
This circumstance makes preprocessing and execution unnecessarily complex. 
Gilbert offers a fixpoint operator \code{fixpoint}, which iteratively applies a given update function $f$ on the previous result of $f$, starting with an initial value $x$ at iteration $0$.
\begin{displaymath}
  n^{th}\text{ iteration}\equiv\underbrace{f(f(\ldots(f(x))\ldots))}_{\text{$n$ times}}
\end{displaymath}
In order to terminate the fixpoint operation, the operator provides two mechanisms.
The user has to specify a maximum number \code{m} of iterations.
Additionally, the user can provide a convergence function \code{c} to the fixpoint operator.
The convergence function is called with the previous and current fixpoint value and returns a boolean value.
Thus, the fixpoint operator terminates either if convergence was detected or if the maximum number of iterations is exceeded.
\begin{equation}
\label{eq:fixpoint}
fixpoint: \underbrace{T}_{\text{\code{x}}} \times \left( \underbrace{T \rightarrow T}_{\text{\code{f}}} \right) \times \underbrace{\mathbb{N}}_{\text{\code{m}}} \times \left(\underbrace{T\times T \rightarrow \mathbb{B}}_{\text{\code{c}}} \right) \rightarrow T
\end{equation}
with $T$ being a generic type variable.

In fact, the fixpoint operator replaces iterations by recursions with a pure update function $f$. 
At this point Gilbert breaks with existing MATLAB code. 
However, all MATLAB loops can be expressed via the fixpoint operator by passing the loop's closure to the update function, see \cref{fig:for2Fixpoint}.
\begin{listing}
  \centering
  \begin{sublisting}{.4\linewidth}
    \begin{lstlisting}[language=Matlab,
      commentstyle=\color{black},
      stringstyle=\color{black},
      keywordstyle=\color{black}\bfseries,
      basicstyle=\footnotesize
    ]
A = 0;
for i = 1:10
  A = A + i;
end
    \end{lstlisting}
    \caption{For loop}
    \label[listing]{fig:for2Fixpoint:for}
  \end{sublisting}
  \begin{sublisting}{.5\linewidth}
      \begin{lstlisting}[language=Matlab, basicstyle=\footnotesize]
f = @(x) ...
  {x{1} + x{2}, x{2} + 1};
r = fixpoint({0,1}, f, 10);
A = r{1};
    \end{lstlisting}
    \caption{Fixpoint}
    \label[listing]{fig:for2Fixpoint:fixpoint}
  \end{sublisting}
  \caption{Transformation from MATLAB for loop \subref{fig:for2Fixpoint:for} to Gilbert fixpoint \subref{fig:for2Fixpoint:fixpoint} formulation. Essentially, all iteration data is combined and passed as a cell array value to the update function.}
  \label{fig:for2Fixpoint}
\end{listing}

@@

MATLAB belongs to the class of dynamically typed languages.
However, the parallel data processing systems, used to run Gilbert programs, have to know the occurring data types.
Therefore, the MATLAB language has to be enriched with type information. 
We infer this type information using the Hindley-Milner (HM) type system~\cite{hindley:tams1969a,milner:jcss1978a} and a slightly derived form of algorithm W~\cite{damas:1982a} for type inference. 
In case that the type inference algorithm cannot properly infer the types, there has to be a way to resolve this problem. 
Similar to \cite{furr:2009a}, we allow to add type information via special comments.

\textbf{Function Overloading}. MATLAB's basic operators, such as \code{+}, \code{-}, \code{/} and \code{*}, for example, are overloaded.
They can be applied to matrices, scalars as well as mixture of both data types.
That makes it very convenient to express mathematical problems, but from a programmer's point of view it causes some hardships.
Originally, HM cannot deal with overloaded functions, because it assumes that each function has a unique type.
In order to extend HM's capabilities, we allowed each function symbol to have a list of signatures.
In the case of \code{+}, the list of signatures would consist of 
\begin{alignat*}{2}
matrix[double] \times matrix[double] &\rightarrow matrix[double]\\
matrix[double] \times double &\rightarrow matrix[double]\\
double \times matrix[double] &\rightarrow matrix[double]\\
double \times double &\rightarrow double
\end{alignat*}
In order to solve the typing problem, the inference algorithm has to resolve this ambiguity.
Having complete knowledge of the argument types is enough to select the appropriate signature.
Sometimes even partial knowledge is sufficient.

@@

Gilbert's matrix type is defined as 
\begin{displaymath}
MatrixType(\underbrace{\tau}_{\text{Element type}},\underbrace{\nu}_{\text{Number of rows}},\underbrace{\nu}_{\text{Number of columns}})
\end{displaymath}
with $\nu$ being the value type. The value type can either be a value variable or a concrete value.

Value variables always appear if we know that a certain relationship holds, but do not know the concrete values yet.
For example, when we multiply two matrices $A\in\mathbb{R}^{a\times b}$ and $B\in\mathbb{R}^{b\times c}$ we know that the result is $A\times B = C \in \mathbb{R}^{a\times c}$, without having knowledge about the actual values of $a$ and $c$.

@@

\textbf{Creation Operators}. Creation operators generate or load some data.
The \code{load} operator takes a path to a stored matrix on disk and the corresponding number of rows and columns as input parameters.
It then reads the stored data and loads the matrix into the program context.
The \code{eye} operator generates an identity matrix of the requested size.
\code{zeros} generates a matrix of the given size which is initialized with zeros.
\code{Randn} takes the size of the resulting matrix and the mean and the standard deviation of a Gauss distribution.
The Gauss distribution is then used to initialize the matrix randomly.

\textbf{Transformation Operators}. The transformation operators constitute the main abstraction of the linear algebra operations.
They group operations with similar properties and thus allow an easier reasoning and optimization of the underlying program. 
The \code{UnaryScalar\-Transformation} takes a single scalar value and applies an unary operation on it. 
The \code{ScalarScalarTransformation} constitutes a binary operation on scalar values whereas \code{ScalarMatrixTransformation} and \code{MatrixScalarTransformation} represent a binary operation between a scalar and a matrix value. 
The \code{VectorwiseMatrixTrans\-formation} applies an operation on each row vector of the given matrix.
A vectorwise operation produces a scalar value for each row vector. 
Similar to the vectorwise operations, the \code{AggregateMatrixTransformation} applies an operation to all matrix entries producing a single scalar result value. 
The iteration mechanism is represented by the \code{Fixpoint\-IterationMatrix} and \code{FixpointIterationCellArray} operators. 
The former operator is used for a fixpoint operation on matrices and the latter on cell arrays. 
Both operators follow the semantics and have the same parameters as the fixpoint abstraction presented in \cref{eq:fixpoint}.

@@

\textbf{Transpose Pushdown}. Transpose pushdown tries to move the transpose operations as close to the matrix input as possible.
Thereby, consecutive transpose operations accumulate at the inputs and unnecessary operations erase themselves.
Consider, for example, $C = A^TB$ and $E = (CD^T)^T$.
By inserting $C$ into $E$, the expression $E=(A^T BD^T)^T$ is obtained. 
This term is equivalent to $DB^T A$. 
The latter formulation contains only one transpose operation. 
Usually multiple transpose operations occur because they are written for convenience reasons at various places in the code. 
Moreover, in complex programs it is possible that the programmer loses track of them or simply is unaware of the optimization potential. 
Therefore, transpose pushdown can be a beneficial optimization.

>>>>

Gilbert's language has been strongly inspired by MATLAB, which should ease its adoption.
The language is a fully functional subset of MATLAB and its grammar is specified in \cite{Rohrmann2014}.
The elementary data type is the matrix. 
Gilbert supports arbitrary $2$-dimensional matrices whose elements can be \emph{double} or \emph{boolean}.
Vectors are not represented by a special type but instead are treated as a single column/row matrix.
Additionally, scalar \emph{double}, \emph{boolean}, and \emph{string} values are supported. 
Gilbert also implements cell array types.
A cell array is defined using curly braces and commas to separate individual values. 
The cells can be accessed by an index appended in curly braces to the cell array variable. 
\Cref{lst:cellArray} shows how to define and access them.
\begin{listing}[!h]
    \begin{lstlisting}[language=Matlab,
        commentstyle=\color{black},
        stringstyle=\color{black},
        xleftmargin=.2\textwidth,
        basicstyle=\small
    ]
c = {true, 2*2, 'cell', 'array'};
b = c{1} & false; % = false
d = c{2} ^ 2; % = 16
s = {c{4}, c{3}}; % = {'array', 'cell'} 
    \end{lstlisting}
  \caption{Cell array usage in Gilbert. Definition of a 4 element cell array, which is accessed subsequently.}
  \label{lst:cellArray}
\end{listing}

Gilbert supports the basic linear algebra operations defined on matrices and scalars. 
They include among others \code{+}, \code{-}, \code{/}, and \code{*}, whereas \code{*} denotes the matrix-matrix multiplication and all other operations are interpreted cell-wise. 
The cell-wise multiplication is indicated by a point preceding the operator. 
Gilbert also supports comparisons operators such as \code{>}, \code{>=}, \code{==}, and \code{\textasciitilde=}. 
Besides the basic arithmetic operations, the user can also define named functions and anonymous functions. 
The syntax of anonymous functions adheres to the MATLAB syntax: \code{@(x,y) x*x + y*y}.

An important aspect of MATLAB are loops. 
MATLAB permits the user to express \code{for} and \code{while} loops, which can have side effects. 
Parallelizing iterations with side effects is difficult because the referenced external state has to be maintained. 
This circumstance makes preprocessing and execution unnecessarily complex. 
Gilbert offers a fixpoint operator \code{fixpoint}, which iteratively applies a given update function $f$ on the previous result of $f$, starting with an initial value $x$ at iteration $0$.
\small
\begin{displaymath}
  n^{th}\text{ iteration}\equiv\underbrace{f(f(\ldots(f(x))\ldots))}_{\text{$n$ times}}
\end{displaymath}
\normalsize

In order to terminate the fixpoint operation, the operator provides two mechanisms.
The user has to specify a maximum number \code{m} of iterations.
Additionally, the user can provide a convergence function \code{c} to the fixpoint operator.
The convergence function is called with the previous and current fixpoint value and returns a boolean value.
Thus, the fixpoint operator terminates either if convergence was detected or if the maximum number of iterations is exceeded.
\small
\begin{equation}
\label{eq:fixpoint}
fixpoint: \underbrace{T}_{\text{\code{x}}} \times \left( \underbrace{T \rightarrow T}_{\text{\code{f}}} \right) \times \underbrace{\mathbb{N}}_{\text{\code{m}}} \times \left(\underbrace{T\times T \rightarrow \mathbb{B}}_{\text{\code{c}}} \right) \rightarrow T
\end{equation}
\normalsize
with $T$ being a generic type variable.

In fact, the fixpoint operator replaces iterations by recursions with a pure update function $f$. 
At this point Gilbert breaks with existing MATLAB code. 
However, all MATLAB loops can be expressed via the fixpoint operator by passing the loop's closure to the update function, see \cref{fig:for2Fixpoint}.
\begin{listing}[!h]
  \centering
  \begin{sublisting}{.4\linewidth}
    \begin{lstlisting}[language=Matlab,
      commentstyle=\color{black},
      stringstyle=\color{black},
      keywordstyle=\color{black}\bfseries,
      basicstyle=\small
    ]
A = 0;
for i = 1:10
  A = A + i;
end
    \end{lstlisting}
    \caption{For loop}
    \label[listing]{fig:for2Fixpoint:for}
  \end{sublisting}
  \begin{sublisting}{.5\linewidth}
      \begin{lstlisting}[language=Matlab, basicstyle=\footnotesize]
f = @(x) ...
  {x{1} + x{2}, x{2} + 1};
r = fixpoint({0,1}, f, 10);
A = r{1};
    \end{lstlisting}
    \caption{Fixpoint}
    \label[listing]{fig:for2Fixpoint:fixpoint}
  \end{sublisting}
  \caption{Transformation from MATLAB for loop \subref{fig:for2Fixpoint:for} to Gilbert fixpoint \subref{fig:for2Fixpoint:fixpoint} formulation. Essentially, all iteration data is combined and passed as a cell array value to the update function.}
  \label{fig:for2Fixpoint}
\end{listing}

@@

MATLAB belongs to the class of dynamically typed languages.
However, the parallel data processing systems used to run Gilbert programs need to know the occurring data types.
Therefore, the MATLAB language has to be enriched with type information. 
We infer this type information using the Hindley-Milner (HM) type system~\cite{hindley:tams1969a,milner:jcss1978a} and a slightly derived form of algorithm W~\cite{damas:1982a} for type inference. 
In case that the type inference algorithm cannot properly infer the types, there has to be a way to resolve this problem. 
Similar to \cite{furr:2009a}, we allow to add type information via special comments.

\textbf{Function Overloading}. MATLAB's basic operators, such as \code{+}, \code{-}, \code{/} and \code{*}, for example, are overloaded.
They can be applied to matrices, scalars as well as mixture of both data types.
That makes it very convenient to express mathematical problems, but from a programmer's point of view it causes some hardships.
Originally, HM cannot deal with overloaded functions, because it assumes that each function has a unique type.
In order to extend HM's capabilities, we allowed each function symbol to have a list of signatures.
In the case of \code{+}, the list of signatures would consist of 
\small
\begin{alignat*}{2}
matrix[double] \times matrix[double] &\rightarrow matrix[double]\\
matrix[double] \times double &\rightarrow matrix[double]\\
double \times matrix[double] &\rightarrow matrix[double]\\
double \times double &\rightarrow double
\end{alignat*}
\normalsize

In order to solve the typing problem, the inference algorithm has to resolve this ambiguity.
Having complete knowledge of the argument types is enough to select the appropriate signature.
Sometimes even partial knowledge is sufficient.

@@

Gilbert's matrix type is defined as 
\small
\begin{displaymath}
MatrixType(\underbrace{\tau}_{\text{Element type}},\underbrace{\nu}_{\text{Number of rows}},\underbrace{\nu}_{\text{Number of columns}})
\end{displaymath}
\normalsize
with $\nu$ being the value type. The value type can either be a value variable or a concrete value.

@@

\textbf{Creation Operators}. Creation operators generate or load some data.
The \code{load} operator loads a matrix from disk with the given number of rows and columns.
The \code{eye} operator generates an identity matrix of the requested size.
\code{zeros} generates a matrix of the given size which is initialized with zeros.
\code{Randn} takes the size of the resulting matrix and the mean and the standard deviation of a Gauss distribution.
The Gauss distribution is then used to initialize the matrix randomly.

\textbf{Transformation Operators}. The transformation operators constitute the main abstraction of the linear algebra operations.
They group operations with similar properties and thus allow an easier reasoning and optimization of the underlying program. 
The \code{UnaryScalar\-Transformation} takes a single scalar value and applies an unary operation on it. 
The \code{ScalarScalarTransformation} constitutes a binary operation on scalar values whereas \code{ScalarMatrixTransformation} and \code{MatrixScalarTransformation} represent a binary operation between a scalar and a matrix value. 
The \code{VectorwiseMatrixTrans\-formation} applies an operation on each row vector of the given matrix.
A vectorwise operation produces a scalar value for each row vector. 
The \code{AggregateMatrixTransformation} applies an operation to all matrix entries producing a single scalar result value. 
The iteration mechanism is represented by the \code{Fixpoint\-IterationMatrix} and \code{FixpointIteration\-CellArray} operators.

@@

\textbf{Transpose Pushdown}. Transpose pushdown tries to move the transpose operations as close to the matrix input as possible.
Thereby, consecutive transpose operations accumulate at the inputs and unnecessary operations erase themselves.
Consider, for example, $C = A^TB$ and $E = (CD^T)^T$.
By inserting $C$ into $E$, the expression $E=(A^T BD^T)^T$ is obtained which is equivalent to $DB^T A$. 
The latter formulation contains only one transpose operation. 
Usually multiple transpose operations occur because they are written for convenience reasons at various places in the code. 
Moreover, in complex programs it is possible that the programmer loses track of them or simply is unaware of the optimization potential. 
Therefore, transpose pushdown can be a beneficial optimization.
